Effects of epidemiological structure on the transient evolution of HIV virulence

### 2016-10-17 ms/plos_latex.tex
still working ...

@@

\end{flushleft}
% Please keep the abstract below 300 words
% current is 298!
\section*{Abstract}
The evolutionary dynamics of parasite virulence over the
  course of an emerging epidemic have important implications both for
  our basic understanding of epidemiological dynamics and,
  potentially, for the outcomes of public health interventions.
  Changes in the fitness landscape
  will generally select for higher virulence during the early
  phase of an epidemic, but quantitative outcomes
  may depend sensitively on biological details and the structure of
  mathematical models used to capture them.  Fraser and
  co-workers have proposed a series of models for eco-evolutionary
  dynamics of HIV that are relatively detailed in their portrayal of
  the tradeoffs between transmission and virulence (mediated by
  set-point viral load, SPVL) and their heritability between
  hosts. However, these models use implicit
  representations of the transmission process that drastically simplify the
  partnership dynamics that previous research has found to be critical
  in driving epidemics of sexually transmitted diseases.  We
  explore models that combine HIV virulence tradeoffs with a range of
  epidemiological structures, modeling partnership formation and
  dissolution and allowing for individuals to transmit disease outside
  of partnerships. We assess summary statistics such as the minimum
  time of progression to AIDS and the time at which this 
  extreme value (peak virulence) occurs across all
  models for a range of 
  partnership dynamic parameters 
  applicable to the HIV epidemic in sub-Saharan Africa. In
  order to account for the different interpretations of parameters
  across model structures, we scale all parameter sets to constrain
  the simulated epidemic growth rate to be identical, matching a
  realistic baseline value. 
  Although virulence trajectories are broadly similar
  across model structures --- an initial increase in virulence, 
  followed by a peak and a decline to an equilibrium virulence slightly
  lower than a peak --- the timing and magnitude of the peak vary
  considerably.
  Models of intermediate
  complexity predicted much lower peak virulence (15 years to
  progress to AIDS) compared to both more realistic models
  and simple random-mixing models with no partnership structure
  at all (both approx. 7.25 years to progress to AIDS).
  In this range of models, the simplest random-mixing structure is
  actually the best approximation to the most realistic model; this
  surprising outcome occurs because the dominance of extra-pair
  contact in the realistic model tends to mask the effects of
  partnership structure.

@@

As we discuss in detail below, the
existing models of HIV eco-evolutionary dynamics either use implicit
models that incorporate the average effects of within-couple sexual
contact --- without representing the explicit dynamics of pair
formation and dissolution or accounting for extra-partnership contact
--- or use an agent-based formulation with parameters that effectively
lead to random mixing among infected and uninfected individuals. Here
we explore the effects of incorporating \emph{explicit}
epidemiological structure in eco-evolutionary models.

We add complexity to the epidemiological model following the general
approach of Champredon \etal\ \cite{champredon_hiv_2013},
which is in turn based on work of Dietz and Hadeler
\cite{dietz_epidemiological_1988}; individuals
join and leave partnerships at a specified rate, and can have sexual
contact both within and outside of established partnerships. In order
to explore how virulence evolution depends on epidemiological
structure, we consider a series of models with increasing levels of
complexity. In order to avoid dependence of the results on a
particular set of parameters --- as we explain below, finding matching
sets of parameters across models with widely differing epidemiological
structures is challenging --- we evaluate our models across a wide
range of parameters, again following Champredon
\etal\ \cite{champredon_hiv_2013} in using a Latin hypercube
design. For each model run, we compute a set of metrics (peak
virulence, timing of virulence peak, equilibrium virulence) that
summarize the evolutionary trajectory of a simulated HIV epidemic.

@@

\begin{equation}
\begin{split}
D_A(\alpha) &= \frac{\tsub{D}{max} D_{50} ^{D_k}}{V_\alpha ^{D_k} + D_{50}^{D_k}}, \\
\beta_A(\alpha) &= \frac{\tsub{\beta}{max} V_\alpha ^ {\beta_k}}%
{V_\alpha^{\beta_k} + \beta_{50} ^{\beta_k}},
\end{split}
\end{equation}
where $V_{\alpha} = 10^\alpha$. 
The uncoupled and extra-couple transmission rates are scaled by multiplying the within-couple transmission rate $\beta$ by the contact ratios $c_u/c_w$ and $c_e/c_w$ (see Appendix). Although such simplification could certainly affect the evolution of virulence (Kret\cite{kretzschmar_effect_1998} , our simplified model is able to produce qualitatively similar results that is qualitatively similar to that of Shirreff \etal's \cite{shirreff_transmission_2011} (\figurename~\ref{fig:panel3}). As our focus is not making accurate predictions, we believe that being able to produce qualitatively similar dynamics is sufficient for this study.

\begin{figure}[!ht]
\includegraphics[width=\textwidth]{../figures/fig1.pdf}
\caption{{\bf Baseline dynamics.}
Time series of mean population \Lspvl\. (a) Shirreff model, effects of varying exponential growth rate, $r$. (b) Effects of varying initial infectious density $I(0)$. (c) Effects of varying initial mean virulence $\alpha(0)$. Orange curve in panel (a) shows that our simplified model is able to produce similar trajectory compared to Shirreff \etal's model using the same parameter set \cite{shirreff_transmission_2011}. In order to compare the models under a similar environment, we match the initial exponential growth rate (see main text "Simulation runs"). Furthermore, we experiment with wide ranges of initial values to show that evolutionary dynamics are sensitive to starting parameters but maintains qualitative features.}
\label{fig:panel3}
\end{figure}

@@

Like Shirreff \etal\ \cite{shirreff_transmission_2011} we incorporate a between-host mutation process in the SPVL, but simplify Shirreff \etal's evolutionary model by using a one-to-one genotype-phenotype mapping.
The mutational process in our model is directly taken from Shirreff \etal. Over the course of infection, mutation occurs within the host. However, it is assumed that SPVL of an infected individual is determined by the SPVL at the time of infection (and is not further affected by within-host mutation). Instead, the mutational effect takes place when an infected individual transmits the virus to a susceptible individual. First, the distribution of \Lspvl\ is discretized into a vector:
\begin{equation}
\alpha_i = \tsub{\alpha}{min} + (\tsub{\alpha}{max} - \tsub{\alpha}{min})\frac{i-1}{n-1} \qquad i = 1,2,3, \dots n.
\end{equation}
We have experimented with varying degrees of discretization in the strain distribution (i.e., values of $n$); in our model runs comparing results with Shirreff \etal\ \cite{shirreff_transmission_2011} (\figurename~\ref{fig:panel3}) we use $n=51$ (i.e. a bin width of 0.1 \Lspvl\ for $\alpha$), but we find only small differences when reducing $n$ to 21 (bin width = 0.25 \Lspvl), which we use for all other simulations.

@@

We developed seven multi-strain evolutionary models covering a gamut between Champredon \etal's relatively realistic \cite{champredon_hiv_2013} and Shirreff \etal's relatively simple \cite{shirreff_transmission_2011} epidemiological structures, each of which is based on different assumptions regarding contact structure and partnership dynamics. Specifically, we focus on the effects of the assumptions of (1) instantaneous vs. non-instantaneous partnership formation, (2) zero vs. positive extra-partnership sexual contact and transmission and (3) homogeneous vs. heterogeneous mixing on the evolution of mean \Lspvl.

Our first four models consider explicit partnership dynamics and are based on Champredon \etal's model \cite{champredon_hiv_2013}. The first two (``pair-formation'' or ``pairform'' for short) assume non-instantaneous partnership formation (i.e. individuals spend some time uncoupled, outside of partnerships) and consist of five states that are classified by infection status and partnership status. $S$ is the number of single (uncoupled) susceptible individuals, and $I$ is the number of single infected individuals. $SS$ is the number of concordant negative (susceptible-susceptible) couples, $SI$ is the number of serodiscordant (susceptible-infected) couples, and $II$ is the number of concordant positive (infected-infected) couples. The first (``pairform+epc'') includes extra-partnership contact (with both uncoupled individuals and individuals in other partnerships) whereas the second (``pairform'') only considers within-couple transmission. The next two models assume instantaneous partnership formation (``instswitch'') and thus consist of only the three partnered states: $SS$, $SI$, and $II$. Like the first two models, these models differ in their inclusion of extra-pair contact: the third model (``instswitch+epc'') includes extra-partnership contact (now only with individuals in other partnerships, since uncoupled individuals don't exist in this model) and the fourth (``instswitch'') only considers within-couple transmission.

The next two models do not explicitly track sexual partnerships.  One (``implicit'') is an implicit serial monogamy model based on the epidemiological model used by Shirreff \etal\ \cite{shirreff_transmission_2011}. It is actually a random mixing model that consist of only two states, $S$ and $I$, and does not consider explicit partnership dynamics. However, to reflect the effect of (instantaneously formed) partnership structure, it uses an adjusted transmission rate that is derived from an approximation of the basic reproduction number of a serial monogamy model \cite{hollingsworth_hiv1_2008}. Another model (``random'') is a simple random-mixing model.

Lastly, we add heterogeneous mixing to pairform+epc model (``hetero''). Individuals are divided into different risk groups based on the sexual activity level, and we assume that sexual activity level is directly proportional to number of non-cohabiting (extra-couple and uncoupled) partners per year \cite{omori2015dynamics}. Model details can be found in the appendix.

@@

\begin{table}[h!]
\caption{Parameter ranges/values.  Note that $c$ and $\rho$ values are doubled from Champredon \etal\  because we keep track of individuals, while they keep track of couples. Starred (*) parameters (used in \figurename~\ref{fig:panel3}), and descriptions of Hill function coefficients, are taken from \cite{shirreff_transmission_2011}.}
\centering
\begin{tabular}{c p{2in} c l}
\hline 
Notation & Description & Range/Value & Source\\
\hline % inserts single horizontal line
$\rho$ & Partnership formation rate & 1/10-2/5 per year & \cite{champredon_hiv_2013} \\
$c$ & Partnership dissolution rate & 1/15-1/5 (1.25*) per year & \cite{champredon_hiv_2013} \\
$c_u/c_w$ & Relative contact rate for uncoupled transmission & 1/5-5 & Assumption \\
$c_e/c_w$ & Relative contact rate extra-couple & 0.01-1 & \cite{champredon_hiv_2013} \\
$\beta_P$ & Rate of transmission during primary infection & 1.31-5.09 (2.76*) per year & \cite{hollingsworth_hiv1_2008} \\
$\beta_D$ & Rate of transmission during high transmission disease stage & 0.413-1.28 (0.76*) per year & \cite{hollingsworth_hiv1_2008} \\
$D_P$ & Duration of primary infection & 1.23/12-6/12 (0.25*) years & \cite{hollingsworth_hiv1_2008} \\
$D_D$ & Duration of high transmission disease stage & 4.81/12-14/12 (0.75*) years & \cite{hollingsworth_hiv1_2008} \\
$\tsub{\beta}{max}$ & Maximum rate of transmission during asymptomatic stage & 0.317 per year & \cite{shirreff_transmission_2011} \\
$\beta_{50}$ & SPVL at which infectiousness is half maximum & 13938 copies per ml & \cite{shirreff_transmission_2011} \\
$\beta_k$ & Hill coefficient: steepness of increase in infectiousness as a function of SPVL & 1.02 & \cite{shirreff_transmission_2011} \\
$\tsub{D}{max}$ & Duration of primary infection & 25.4 years & \cite{shirreff_transmission_2011} \\
$D_{50}$ & SPVL at which duration of asymptomatic infection is half maximum & 3058 copies per ml & \cite{shirreff_transmission_2011} \\
$D_{k}$ & Hill coefficient: steepness of decrease in duration as a function of SPVL & 0.41 & \cite{shirreff_transmission_2011} \\
$\sigma_M$ & Mutation standard deviation of $\log_{10}$ SPVL & 0.12 & \cite{shirreff_transmission_2011} \\
$\tsub{\alpha}{min}$ & Minimum $\log_{10}$ SPVL & 2 & \cite{shirreff_transmission_2011}\\
$\tsub{\alpha}{max}$ & Maximum $\log_{10}$ SPVL & 7 & \cite{shirreff_transmission_2011}\\
$n$ & Number of strains & 21 (51*) & Assumption\\
$\mu$ & Mean number of non-cohabiting sexual partners & 0.103 - 1.206 & \cite{omori2015dynamics}\\
$\kappa$ & Squared coefficient of variation of number of non-cohabiting sexual partners & 0.01 - 100 & Assumption\\
\hline
\end{tabular}
\label{table:parmsTable}
\end{table}

@@

@@

For each model we derive the following summary statistics: peak virulence, peak time, equilibrium virulence, and relative peak virulence. The transient phase of an epidemic is often characterized by high virulence, and we define peak virulence as the maximum virulence during this phase. It is simply calculated by taking the maximum value from the virulence trajectory, and peak time is the time at which the maximum value is reached. Once the epidemic enters the endemic phase, evolution of virulence stabilizes and reaches equilibrium. Equilibrium virulence is calculated by taking the mean virulence at 4000 years. Although most simulations reach equilibrium much earlier, we set our time horizon at a much later date as some simulation runs have slow rate of evolution depending on the parameter set and model assumptions.

We focus on these statistics for the following two reasons. First of all, knowing the possible ranges for the peak virulence allows us to estimate the worst-case scenario for the HIV and other sexually transmitted disease epidemics. Pathogens may already have evolved towards high virulence during the early stages of an epidemic, by the time it is observed by public health authorities. Understanding how virulent a pathogen can evolve before an epidemic begins can be helpful for controlling the disease. Furthermore, knowing the initial virulence, peak virulence, timing of the peak virulence, and equilibrium virulence provide sufficient detail to identify the shape of the virulence trajectory. During an epidemic outbreak, it is difficult to observe virulence evolution. Specifically, in the case of HIV and other sexually transmitted diseases, slow evolutionary time-scale makes observing changes in the mean virulence even more challenging. Knowing the ranges of these statistics can help real-time virulence evolution prediction during an epidemic less troublesome.

@@

Across the entire range of parameters covered by the LHS analysis, all of the classes of models we considered produce qualitatively similar virulence trajectories (\figurename~\ref{fig:virtraj}). Although the speed of virulence evolution varies, leading to wide variation in the peak virulence (means ranging from approximately 3.75 to 4.5 \Lspvl), virulence peaks in all models between 200 and 300 years.

\begin{figure}[!ht]
\includegraphics[width=\textwidth]{../figures/fig2.pdf}
\caption{{\bf Envelopes of virulence trajectories under all models.}
All models were run until $t=4000~\textrm{years}$; truncated series are shown here.}
\label{fig:virtraj}
\end{figure}

Our chosen summary statistics (peak time, peak virulence, equilibrium
virulence, and relative peak virulence) all vary considerably across models
\ref{fig:unidist}.
We first consider the models of intermediate realism: implicit,
instantaneous-switching with and without extra-pair contact, and
pair formation without extra-pair contact. Some parameter
sets for these models lead to low equilibrium virulence ($\approx 2.5 \log_{10}$ SPVL);
these same sets lead to correspondingly low
peak virulence ($< 3.5 \log_{10}$ SPVL) and early peak times (before 200 years), 
but high relative peaks ($>1.3$)
(\figurename~\ref{fig:pairplot}, leftmost column) because the equilibrium virulence is low.
At the opposite extreme, parameter sets that produce high equilibrium virulence 
also produce late peaks ($> 200 \text{ years}$), 
high peak virulence, and low relative peaks ($\approx 1.05$).
The pair-formation without extra-pair contact and implicit models
occasionally have parameter sets that select for such low virulence across
the board that they never exceed their initial virulence, leading to a tail
of peak times near zero.

@@

The most striking aspect of the univariate comparisons in
\figurename~\ref{fig:unidist}, (and the bivariate comparisons in
\figurename~\ref{fig:pairplot}) is the similarity between the results of the
least (random mixing) and the most complex (pair formation with
extra-pair contact) models. The random-mixing model has lower
variability, because it is unaffected by uncertainty in pair formation
and extra-pair contact parameters, but otherwise the virulence
dynamics of these two extreme models are remarkably similar.
This phenomenon is driven by the strong effects of extra-pair contact in the
model with explicit pair formation and extra-pair contact 
(``pairform+epc'' in \figurename{}s~\ref{fig:virtraj}-\ref{fig:plot_sens}). When individuals spend time uncoupled between
partnerships, and when these single individuals can transmit disease
to coupled individuals, the resulting unstructured mixing overwhelms
the effect of structured mixing within couples, leading to mixing
that is effectively close to random.

Expressing these results in terms of more directly interpretable epidemic
parameters, i.e. using the Hill functions to translate \Lspvl\ to
within-couple transmission probabilities and average time of progression
to AIDS, shows that these differences are practically as well as 
scientifically important. The random-mixing and pairform+epc models
predict minimum times to progression (at the virulence peak) of
5.7 (95\% CI 6.1-6.3) and 6.0 (5.0-7.7) years respectively, while
the implicit model gives progression times about twice as long:
12.5 (9.5-15.6) years. The corresponding differences in 
within-couple transmission
probability are even more extreme, about a fourfold difference:
0.249 (0.24-0.26) and 0.252 (0.19-0.28) per year for the 
random and pairform+epc models vs. 0.059 (0.02-0.13) per year
for the implicit model (Figs~???? show univariate distributions
on the epidemiological scales
of progression time and transmission probability,
for all summary measures and all models).

@@

% \bibliography{virulence}
\begin{thebibliography}{10}

\bibitem{Ebert1999}
Ebert D.
\newblock The evolution and expression of parasite virulence.
\newblock In: Stearns SC, editor. Evolution in Health \& Disease. New York:
  Oxford University Press, Oxford, UK; 1999. p. 161--172.

\bibitem{EbertBull2003}
Ebert D, Bull JJ.
\newblock Challenging the trade-off model for the evolution of virulence: is
  virulence management feasible?
\newblock Trends Microbiol. 2003;11(1):15--20.

>>>>

\newcommand{\todo}[1]{\textbf{#1}}

@@

\end{flushleft}
% Please keep the abstract below 300 words
% current is 298!
\section*{Abstract}
The evolutionary dynamics of parasite virulence over the
  course of an emerging epidemic have important implications both for
  our basic understanding of epidemiological dynamics and,
  potentially, for the outcomes of public health interventions.
  Changes in the fitness landscape
  will generally select for higher virulence during the early
  phase of an epidemic, but quantitative outcomes
  may depend sensitively on biological details and the structure of
  mathematical models used to capture them.  Fraser and
  co-workers have proposed a series of models for eco-evolutionary
  dynamics of HIV that are relatively detailed in their portrayal of
  the tradeoffs between transmission and virulence (mediated by
  set-point viral load, SPVL) and their heritability between
  hosts. However, these models use implicit
  representations of the transmission process that drastically simplify the
  partnership dynamics that previous research has found to be critical
  in driving epidemics of sexually transmitted diseases.  We
  explore models that combine HIV virulence tradeoffs with a range of
  epidemiological structures, modeling partnership formation and
  dissolution and allowing for individuals to transmit disease outside
  of partnerships. We assess summary statistics such as the minimum
  time of progression to AIDS and the time at which this 
  extreme value (peak virulence) occurs across all
  models for a range of 
  partnership dynamic parameters 
  applicable to the HIV epidemic in sub-Saharan Africa. In
  order to account for the different interpretations of parameters
  across model structures, we scale all parameter sets to constrain
  the simulated epidemic growth rate to be identical, matching a
  realistic baseline value. 
  Although virulence trajectories are broadly similar
  across model structures --- an initial increase in virulence/decrease
  in progression time, followed by a minimum and an increase to an equilibrium 
  progression time slightly higher than the minimum --- the timing and magnitude of the minimum vary
  considerably.
  Models of intermediate
  complexity predicted much lower peak virulence (15 years to
  progress to AIDS) compared to both more realistic models
  and simple random-mixing models with no partnership structure
  at all (both approx. 7.25 years to progress to AIDS).
  In this range of models, the simplest random-mixing structure is
  actually the best approximation to the most realistic model; this
  surprising outcome occurs because the dominance of extra-pair
  contact in the realistic model tends to mask the effects of
  partnership structure.

@@

As we discuss in detail below,
existing models of HIV eco-evolutionary dynamics either use implicit
models that incorporate the average effects of within-couple sexual
contact --- without representing the explicit dynamics of pair
formation and dissolution or accounting for extra-partnership contact
--- or use an agent-based formulation with parameters that effectively
lead to random mixing among infected and uninfected individuals. Here
we explore the effects of incorporating \emph{explicit}
epidemiological structure in eco-evolutionary models.

We add complexity to the epidemiological model following the general
approach of Champredon \etal\ \cite{champredon_hiv_2013},
which is in turn based on work of Dietz and Hadeler
\cite{dietz_epidemiological_1988}; individuals
join and leave partnerships at a specified rate, and can have sexual
contact both within and outside of established partnerships. In order
to explore how virulence evolution depends on epidemiological
structure, we consider a series of models with increasing levels of
complexity. In order to avoid dependence of the results on a
particular set of parameters --- as we explain below, finding matching
sets of parameters across models with widely differing epidemiological
structures is challenging --- we evaluate our models across a wide
range of parameters, again following Champredon
\etal\ \cite{champredon_hiv_2013} in using a Latin hypercube
design. For each model run, we compute a set of metrics (minimum
progression time/peak virulence, timing of maximum virulence, equilibrium virulence) that summarize the evolutionary trajectory of a simulated HIV epidemic.

@@

\begin{equation}
\begin{split}
D_A(\alpha) &= \frac{\tsub{D}{max} D_{50} ^{D_k}}{V_\alpha ^{D_k} + D_{50}^{D_k}}, \\
\beta_A(\alpha) &= \frac{\tsub{\beta}{max} V_\alpha ^ {\beta_k}}%
{V_\alpha^{\beta_k} + \beta_{50} ^{\beta_k}},
\end{split}
\label{eq:hillfuns}
\end{equation}
where $V_{\alpha} = 10^\alpha$. 
The uncoupled and extra-couple transmission rates are scaled by multiplying the within-couple transmission rate $\beta$ by the contact ratios $c_u/c_w$ and $c_e/c_w$ (see Appendix). Simplifying the model of HIV pathogenesis from three stages to a single stage could affect our conclusions about the evolution of virulence (e.g. Kretzschmar and Dietz \cite{kretzschmar_effect_1998} show that pair formation dynamics and multiple stages of infectivity have interactive effects on $\rzero$). However, our simplified model produces results that are qualitatively similar to those of Shirreff \etal's \cite{shirreff_transmission_2011} model; when our model is calibrated to have a similar initial epidemic growth rate $r$, the peak \Lspvl\ occurs at the same time ($\approx$ 200 years) but slightly higher (4.7 vs. 4.3); the relative height of the peak is \todo{compute difference and fill this in}
(\figurename~\ref{fig:panel3}).

\begin{figure}[!ht]
\includegraphics[width=\textwidth]{../figures/fig1.pdf}
\caption{{\bf Baseline dynamics.}
Time series of mean population \Lspvl. (a) Contrast with three-stage Shirreff model and effects of varying exponential growth rate, $r$. (b) Effects of varying initial infectious density $I(0)$. (c) Effects of varying initial mean virulence $\alpha(0)$. The orange curve in panel (a) shows that our simplified model can produce similar trajectories to those Shirreff \etal's model \cite{shirreff_transmission_2011}. In order to compare the models, we match the initial exponential growth rate (see main text ``Simulation runs"). Furthermore, we experiment with wide ranges of initial values (panels b and c) to show that evolutionary dynamics are not overly sensitive to starting parameter values, which we hold constant in our simulations.}
\label{fig:panel3}
\end{figure}

@@

Like Shirreff \etal\ \cite{shirreff_transmission_2011} we incorporate a between-host mutation process in the SPVL. We simplify Shirreff \etal's evolutionary model by using a one-to-one genotype-phenotype mapping rather
than allowing for variation in phenotypes of a single genotype.
The mutational process in our model is directly taken from Shirreff \etal. Over the course of infection, mutation occurs within the host. However, it is assumed that SPVL of the strain transmitted by an infected individual is determined by the SPVL at the time of infection and is not further affected by within-host mutation. Instead, the mutational effect takes place in a single
step at the time of transmission. First, the distribution of \Lspvl\ is discretized into a vector:
\begin{equation}
\alpha_i = \tsub{\alpha}{min} + (\tsub{\alpha}{max} - \tsub{\alpha}{min})\frac{i-1}{n-1} \qquad i = 1,2,3, \dots n.
\end{equation}
We have experimented with varying degrees of discretization in the strain distribution (i.e., values of $n$); in our model runs comparing results with Shirreff \etal\ \cite{shirreff_transmission_2011} (\figurename~\ref{fig:panel3}) we use $n=51$ (i.e. a bin width of 0.1 \Lspvl\ for $\alpha$), but reducing $n$ to 21 (bin width = 0.25 \Lspvl) makes little difference; we use this coarser grid for all other simulations reported.

@@

We developed seven multi-strain evolutionary models covering a gamut between Champredon \etal's relatively realistic \cite{champredon_hiv_2013} and Shirreff \etal's relatively simple \cite{shirreff_transmission_2011} epidemiological structures, each of which is based on different assumptions regarding contact structure and partnership dynamics. Specifically, we focus on the effects of the assumptions of (1) instantaneous vs. non-instantaneous partnership formation, (2) zero vs. positive extra-partnership sexual contact and transmission and (3) homogeneous vs. heterogeneous levels of sexual activity on the evolution of mean \Lspvl.

Our first four models consider explicit partnership dynamics and are based on Champredon \etal's model \cite{champredon_hiv_2013}. The first two (``pair-formation'' or ``pairform'' for short) assume non-instantaneous partnership formation (i.e. individuals spend some time uncoupled, outside of partnerships) and consist of five states that are classified by infection status and partnership status. $S$ is the number of single (uncoupled) susceptible individuals, and $I$ is the number of single infected individuals. $SS$ is the number of concordant negative (susceptible-susceptible) couples, $SI$ is the number of serodiscordant (susceptible-infected) couples, and $II$ is the number of concordant positive (infected-infected) couples. The first (``pairform+epc'') includes extra-partnership contact (with both uncoupled individuals and individuals in other partnerships) whereas the second (``pairform'') only considers within-couple transmission.

The next two models, which are intended to bridge the gap between models with fully explicit pair-formation dynamics and simpler, implicit models, assume instantaneous partnership formation (``instswitch''). The compartmental structure thus omits the single states $S$ and $I$, comprising only the three partnered states: $SS$, $SI$, and $II$. Like the first two models, this pair of models differs in their inclusion of extra-pair contact: the third model (``instswitch+epc'') includes extra-partnership contact (now only with individuals in other partnerships, since uncoupled individuals do not exist in this model) while the fourth (``instswitch'') only considers within-couple transmission.

The next two models represent extreme simplifications of sexual partnership dynamics.  One (``implicit'') is an implicit serial monogamy model based on the epidemiological model used by Shirreff \etal\ \cite{shirreff_transmission_2011}. It is actually a random mixing model that consist of only two states, $S$ and $I$, and does not consider explicit partnership dynamics. However, to reflect the effect of partnership structure, it uses an adjusted transmission rate that is derived from an approximation of the basic reproduction number of a serial monogamy model with instantaneous pair formation \cite{hollingsworth_hiv1_2008}. The second model of this pair (``random'') is a simple random-mixing model.

Lastly, we add a model of heterogeneity in sexual activity to the pairform+epc model (``hetero''). Individuals are divided into different risk groups based on the sexual activity level; we scale all aspects of sexual activity, assuming that sexual activity level in both within- and extra-couple contacts is directly proportional to number of non-cohabiting (extra-couple and uncoupled) partners per year \cite{omori2015dynamics} (see Appendix for model details). We assume random activity-weighted mixing between risk groups \todo{cite old Anderson-May HIV paper?}. While this model lacks some 
important elements, such as age-structured mixing patterns, needed for realistic models of HIV transmission in sub-Saharan Africa, we see it as a first step toward assessing the effects of epidemiological complexity. As even these models are pushing the limits of compartmental-based models, adding further complexity will probably require a shift to an agent-based model framework, as well as considerable effort in model calibration \cite{herbeck_hiv_2014,delva_connecting_2016}.

@@

\begin{table}[h!]
\caption{Parameter ranges/values.  Values of $c$ and $\rho$ are doubled from those given by Champredon \etal; we keep track of individuals in the model, while they keep track of couples. Starred (*) parameters (used in \figurename~\ref{fig:panel3}), and descriptions of Hill function coefficients, are taken from \cite{shirreff_transmission_2011}.}
\centering
\begin{tabular}{c p{2in} c l}
\hline 
Notation & Description & Range/Value & Source\\
\hline % inserts single horizontal line
$\rho$ & Partnership formation rate & 1/10-2/5 per year & \cite{champredon_hiv_2013} \\
$c$ & Partnership dissolution rate & 1/15-1/5 (1.25*) per year & \cite{champredon_hiv_2013} \\
$c_u/c_w$ & Relative contact rate for uncoupled transmission & 1/5-5 & Assumption \\
$c_e/c_w$ & Relative contact rate extra-couple & 0.01-1 & \cite{champredon_hiv_2013} \\
$\beta_P$ & Rate of transmission during primary infection & 1.31-5.09 (2.76*) per year & \cite{hollingsworth_hiv1_2008} \\
$\beta_D$ & Rate of transmission during high transmission disease stage & 0.413-1.28 (0.76*) per year & \cite{hollingsworth_hiv1_2008} \\
$D_P$ & Duration of primary infection & 1.23/12-6/12 (0.25*) years & \cite{hollingsworth_hiv1_2008} \\
$D_D$ & Duration of high transmission disease stage & 4.81/12-14/12 (0.75*) years & \cite{hollingsworth_hiv1_2008} \\
$\tsub{\beta}{max}$ & Maximum rate of transmission during asymptomatic stage & 0.317 per year & \cite{shirreff_transmission_2011} \\
$\beta_{50}$ & SPVL at which infectiousness is half maximum & 13938 copies per ml & \cite{shirreff_transmission_2011} \\
$\beta_k$ & Hill coefficient: steepness of increase in infectiousness as a function of SPVL & 1.02 & \cite{shirreff_transmission_2011} \\
$\tsub{D}{max}$ & Duration of primary infection & 25.4 years & \cite{shirreff_transmission_2011} \\
$D_{50}$ & SPVL at which duration of asymptomatic infection is half maximum & 3058 copies per ml & \cite{shirreff_transmission_2011} \\
$D_{k}$ & Hill coefficient: steepness of decrease in duration as a function of SPVL & 0.41 & \cite{shirreff_transmission_2011} \\
$\sigma_M$ & Mutation standard deviation of $\log_{10}$ SPVL & 0.12 & \cite{shirreff_transmission_2011} \\
$\tsub{\alpha}{min}$ & Minimum $\log_{10}$ SPVL & 2 & \cite{shirreff_transmission_2011}\\
$\tsub{\alpha}{max}$ & Maximum $\log_{10}$ SPVL & 7 & \cite{shirreff_transmission_2011}\\
$n$ & Number of strains & 21 (51*) & Assumption\\
$\mu$ & Mean number of non-cohabiting sexual partners & 0.103 - 1.206 & \cite{omori2015dynamics}\\
$\kappa$ & Squared coefficient of variation of number of non-cohabiting sexual partners & 0.01 - 100 & Assumption\\
\hline
\end{tabular}
\label{table:parmsTable}
\end{table}

@@

\todo{add something about choice of heterogeneity}

@@

Although each disease strain's core characteristic is its SPVL, the
SPVL has one-to-one correspondences (based on eq.~\ref{eq:hillfuns})
with both the expected time to progression to AIDS and with the rate
(probability per unit time) of HIV transmission. Because the time to
progression (measured in years) is easier to interpret than
SPVL (measured in \Lspvl\ units), we summarize the virulence
trajectories for each model run in terms of time to progression
rather than SPVL. Because the  time to progression is inversely
related to SPVL (increasing SPVL decreases the time to progression),
the time to progression is technically measuring inverse
virulence rather than virulence (we did not think that 
reporting virulence as the rate of progression to AIDS, in units
of $\textrm{years}^{-1}$, would help interpretability).
For each model we derive the following summary statistics:
minimum expected time to progression;
time at which this minimum occurs
(corresponding to peak virulence --- this is also the time at which the
maximum rate of progression, maximum SPVL, and maximum transmission rate
occur); equilibrium time to progression; 
and the ratio of progression time at its minimum to the equilibrium
value. Equilibrium progression time is calculated after 4000 years of simulated
time. Although most simulations reach equilibrium much earlier, we set our time horizon at a much later date as some simulation runs have slow rate of evolution depending on the parameter set and model assumptions.

Knowing the minimum progression time, timing of the minimum progression time/peak virulence, and equilibrium progression time provide sufficient detail to identify the overall shape of the virulence trajectory.
In particular, knowing the timing of the peak virulence (how many years
into the epidemic the virulence peaks) can help epidemiologists
guess whether the virulence of an emerging pathogen is likely (1)
to have peaked early, possibly even before the pathogen is detected
spreading in the population, and decline over the remaining course
of the epidemic; (2) to increase, peak, and decline over the
foreseeable future; or (3) to increase very slowly, peaking only
in the far future. To the extent that our simplistic model for HIV
reflects reality, we would take the peak time of 150-300 years 
(Figure~\ref{fig:panel3}c) to mean that, in the absence of
treatment, the epidemic would probably still be increasing in virulence.

@@

Across the entire range of parameters covered by the LHS analysis, all of the classes of models we considered produce qualitatively similar virulence trajectories (\figurename~\ref{fig:virtraj}). Although the speed of virulence evolution varies, leading to wide variation in the minimum progression time (means ranging from approximately \todo{fill in}), virulence peaks in all models between 200 and 300 years.

\begin{figure}[!ht]
\includegraphics[width=\textwidth]{../figures/fig2.pdf}
\caption{{\bf Envelopes of virulence trajectories (expected
time of progression to AIDS) under all models.}
All models were run until $t=4000~\textrm{years}$; truncated series are shown here.}
\label{fig:virtraj}
\end{figure}

Our chosen summary statistics (peak time, minimum progression time, equilibrium
progression time, and relative progression time) all vary considerably across models
\ref{fig:unidist}.
We first consider the models of intermediate realism: implicit,
instantaneous-switching with and without extra-pair contact, and
pair formation without extra-pair contact. Some parameter
sets for these models lead to low equilibrium virulence ($\approx 18$ years
to progression);
these same sets lead to correspondingly low
peak virulence (16 years to progression) and early peak times (before 200 years).
At the opposite extreme, parameter sets that produce high equilibrium virulence (8 years to progression)
also produce late peaks ($> 200~\text{years}$) and
high peak virulence (4 years to progression).
The pair-formation without extra-pair contact and implicit models
occasionally have parameter sets that select for such low virulence across
the board that they never exceed their initial virulence, leading to a tail
of peak times near zero.

@@

The most striking aspect of the univariate comparisons in
\figurename~\ref{fig:unidist}, (and the bivariate comparisons in
\figurename~\ref{fig:pairplot}) is the similarity between the results of the
least (random mixing) and the most complex (pair formation with
extra-pair contact, pairform+epc with heterogeneity) models. The random-mixing model has the lowest variability, because it is unaffected by uncertainty in pair formation and extra-pair contact parameters, but otherwise the virulence
dynamics of these three extreme models are remarkably similar.
This phenomenon is driven by the strong effects of extra-pair contact in the
model with explicit pair formation and extra-pair contact 
(``pairform+epc'' in \figurename{}s~\ref{fig:virtraj}-\ref{fig:plot_sens}). When individuals spend time uncoupled between
partnerships, and when these single individuals can transmit disease
to coupled individuals, the resulting unstructured mixing overwhelms
the effect of structured mixing within couples, leading to mixing
that is effectively close to random.
Once unstructured mixing is strong, adding realistic heterogeneity
of mixing to the model has little effect other than increasing
the variability in the outcomes.

These differences are practically as well as 
scientifically important. The random-mixing and pairform+epc models
predict minimum times to progression (at the virulence peak) of
5.7 (95\% CI 6.1-6.3) and 6.0 (5.0-7.7) years respectively, while
the implicit model gives progression times about twice as long:
12.5 (9.5-15.6) years. The corresponding differences in 
within-couple transmission
probability are even more extreme, about a fourfold difference:
0.249 (0.24-0.26) and 0.252 (0.19-0.28) per year for the 
random and pairform+epc models vs. 0.059 (0.02-0.13) per year
for the implicit model (Figs~???? \todo{fill in!} show univariate distributions
on the epidemiological scales
of progression time and transmission probability,
for all summary measures and all models).

@@

\bibliography{virulence}
\end{document}
